### 2014-11-24 doc/thesis/system.tex
Adding raw text about the refinement of table types

The most unusual feature of our type system is the refinement of
table types, which is given by the following rule:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique} \;\;\;
       \env_{1} \vdash x:t_{1}, \env_{2} \;\;\;
       \not \exists i \in 1..n \; t_{1} \lesssim k_{i}}
      {\env_{1} \vdash n[x] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
\end{array}
\]

@@

@@

Table type as hash.
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``foo"] = 1 \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; x:\Integer \cup \Nil = t[``foo"] \; \mathbf{in} \; x = t[``bar"]$}
\end{tabular}
\end{center}

Table type as array.
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{\Integer:\String \cup \Nil\}_{closed} = \{ ``a", ``b", ``c" \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; x:\String \cup \Nil = t[1] \; \mathbf{in} \; x = t[4]$}
\end{tabular}
\end{center}

Table type as record (first example).
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``l":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``l"] = ``Reed", [``f"] = ``Lou" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}

Table type as record (second example).
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``m":\String \cup \Nil, ``l":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``f"] = ``Lewis", [``m"] = ``Allan", [``l"] = ``Reed" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}

Table type as record (third example).
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``m":\String \cup \Nil, ``l":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``f"] = ``Lou", [``l"] = ``Reed" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}

Refinement of table types (second example).
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; t:\{\}_{open} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = p \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$t[``f"] \; {<}\String{>} = ``Lou";$}\\
& & \multicolumn{1}{l}{$t[``l"] \; {<}\String{>} = ``Reed";$}\\
\end{tabular}
\end{center}

Objects.
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x:"\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
\multicolumn{4}{l}{$\mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:number, y:number):\Self$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \}$}\\
& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:number, y:number):\Void$}\\
& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
\end{tabular}
\end{center}

>>>>

The most unusual feature of our type system is the refinement of
table types:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique} \;\;\;
       \env_{1} \vdash x:t_{1}, \env_{2} \;\;\;
       \not \exists i \in 1..n \; t_{1} \lesssim k_{i}}
      {\env_{1} \vdash n[x] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
\end{array}
\]

@@

This example also uses the rule \textsc{T-CONSTRUCTOR1} to
infer the type of the empty table constructor:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

Typed Lua also has inference rules to infer table types that we
can use to initialize hashes, arrays, and records:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env \vdash c:k_{i}{:}v_{i}, \env_{i} \;\;\;
       n = |\;\vec{c}\;| \;\;\;
       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
      {\env \vdash \{\;\vec{c}\;\}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}, \env}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-CONSTRUCTOR3}\\
\dfrac{\env \vdash {...}:t, \env}
      {\env \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-CONSTRUCTOR4}\\
\dfrac{\begin{array}{c}
       \env \vdash c:k_{i}{:}v_{i}, \env_{i} \;\;\;
       \env \vdash {...}:t, \env \;\;\;
       n = |\;\vec{c}\;| \\
       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t \cup \Nil\}_{unique})
       \end{array}}
      {\env \vdash \{\;\vec{c},\;{...}_{m}\;\}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}v \cup \Nil\}_{unique}, \env}
\end{array}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR2} type checks the following example:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``x"] = 1, [``y"] = 2 \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; b:\Integer \cup \Nil = a[``x"] \; \mathbf{in} \; x = a[``z"]$}
\end{tabular}
\end{center}

It also uses the rules \textsc{T-FIELD1} and \textsc{T-FIELD2}:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-FIELD1}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Boolean \cup \Number \cup \String}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}:super(t_{2}), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD2}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       \Nil \not\subtype t_{1}}
      {\env_{1} \vdash [e_{1}] = e_{2}: \Any:super(t_{2}), \env_{3}}
\end{array}
\end{array}
\]

The rule \textsc{T-FIELD1} infers fields that keys are subype
of $\Boolean$, $\Number$, or $\String$.
The rule \textsc{T-FIELD2} infers fields thaty keys have the
dynamic type if its type does not include $\Nil$.

Typed Lua handles arrays as hashes that maps $\Integer$ to some
type $t$.
We did not include in the abstract syntax tree the simplified
syntax of table consturctors that initializes arrays because
we can write it using explicit indices.
For instance, we write the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to mean the table constructor
$\{ ``x", ``y", ``z" \}$.
We chose to make this simplification because the inference rules
for the simpflified version does not include anything new.
They are the same rules that just automatically control the indices.
For the same reason, we also omitted in the table constructor expressions
that return multiple results.
However, we did not omit the vararg expression in the last position
to show the case that we infer a record with an array part.

The rule \textsc{T-CONSTRUCTOR3} type checks the following example:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\Integer:\String \cup \Nil\}_{closed} = \{ {...}_{m} \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; b:\String \cup \Nil = t[1] \; \mathbf{in} \; b = t[5]$}
\end{tabular}
\end{center}

We are assuming that $...$ is in the environment as has type $\String$.

The rule \textsc{T-CONSTRUCTOR4} type checks the following example:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\Integer, ``y":\Boolean, \Integer:\String \cup \Nil\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m} \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; a[``x"] = ``bar"$}
\end{tabular}
\end{center}

We are also assuming that $...$ is in the environment as has type $\String$.

These rules along \emph{unique} table types also allow us initialize
record types that have optional fields.
We can include the optional field in the initialization
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = ``bar", [``z"] = ``baz" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}
and we can also ommit it
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``z"] = ``baz" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}

Typed Lua also keeps track whether it is safe or not safe to
change the type of a table.
To do that, our type system always promotes an \emph{open}
table to \emph{closed} and an \emph{unique} table to \emph{open}
when they appear as expressions.
The order also matters, as going from \emph{unique} to \emph{open}
before going from \emph{open} to \emph{closed} closes \emph{unique} tables.
The rule \textsc{T-IDREAD} along with the predicate \emph{close}
express this intuition:
\[
\begin{array}{c}
\mylabel{T-IDREAD}\\
\dfrac{\env_{1}(n_{r}) = t}
      {\env_{1} \vdash n_{r}:close(t), \env_{1}}
\end{array}
\]

The rule \textsc{T-IDREAD} makes this unsafe example not type check:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``x":\String \cup \Nil\}_{closed} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String \}_{unique} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a = b \;;\; a[``x"] = \mathbf{nil}$}
\end{tabular}
\end{center}

The assignment $a = b$ is not valid because the aliasing changed the
type of $b$ from $\{``x":\String\}_{unique}$ to $\{``x":\String\}_{open}$,
which is not subtype of $\{``x":\String \cup \Nil\}_{closed}$.

However, sometimes we want to use \emph{unique} tables to initialize
closed tables, as they allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allow us
to use the \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{t \subtype \env_{1}(n)}
      {\env_{1} \vdash {<}t{>} \; n:t, \env_{1}[n \mapsto t]}
\end{array}
\]

The rule \textsc{T-CAST} allow us to type check the following example:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
& & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}

We can use $a$ to initialize $b$ because the cast expression converts
the type of $a$ from $\{``x":\String, ``y":\String\}_{unique}$ to
$\{``x":\String, ``y":\String \cup \Nil\}_{open}$, which is subtype of
$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$.
We can continue to refine the type of $a$ after the aliasing,
as it still holds an \emph{open} table.
At the end of this example, $a$ has type
$\{``x":\String, ``y":\String \cup \Nil, ``z":\Integer\}_{open}$.

Refinement of table types (second example).
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}

Typed Lua uses the refinement of table types to express an object-oriented
idioms that is commonly used by Lua programmers.
\[
\begin{array}{c}
\mylabel{T-METHOD3}\\
\dfrac{\begin{array}{c}
       \env_{1}(n_{1}) = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{unique|open} \;\;\;
       \env_{1} \vdash n_{2} : l \;\;\;
       \not \exists i \in 1..n \; l \lesssim k_{i}\\
       closeall(\env_{1}[self \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed}, \vec{n} \mapsto \vec{t}, \ret \mapsto r]) \vdash s:\env_{2}
       \end{array}}
      {\begin{array}{c}
       \env_{1} \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s:\\
       closeset(\env_{1}[n_{1} \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \rightarrow r\}_{unique|closed}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
       \end{array}}
\end{array}
\]

\[
\begin{array}{c}
\mylabel{T-INVOKE1}\\
\dfrac{\env_{1} \vdash e[n]:p_{1} \rightarrow r, \env_{2} \;\;\;
       \env_{2} \vdash el:p_{2}, \env_{3} \;\;\;
       \Self \times p_{2} \lesssim p_{1}}
      {\env_{1} \vdash e{:}n(el)_{m}:r, \env_{3}}
\end{array}
\]

\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
\multicolumn{4}{l}{$\mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:number, y:number):\Self$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:number, y:number):\Void$}\\
& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
\end{tabular}
\end{center}

@@
